Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-04-20 demystify.tex
updates

Many students lament that they get no interviews or are rejected (with strong profiles) and that the process seems random.  However, while it is true that the process is not perfect, it is not random.
There are many possible reasons!  Other than not having a strong profile (e.g., research potential, GPA, LoRs, SOP, poor interviews), here are some other reasons.

@@

\paragraph{Red flags in Your Background} Plagiarism, cheating, or other academic dishonesty are \red{red} flags. Another one is if you have a history of jumping from one program to another (e.g., you have been in multiple different programs in a short time). How would the reviewers know?  They might have contacts in other programs and heard about you, your LoRs might mention it.

@@

%\section{How exactly do you evaluate an application?}\label{sec:ievaluate}

% Around 15 minutes on average (less than 10 mins for definite rejection; 20+ mins for acceptance). In earlier years I spent more time, but now I can quickly tell if an application is good or bad.

% The evaluation system allows for downloading a single pdf that consists of a summary page (e.g., schools, GPA, standard test scores), transcripts, test results, LoRs, CV, SOP, and writing samples (if any). I typically review the application in this order.  I also make notes on what I've seen, using them to support my evaluation.

% I first skim the summary for school, GPA, standard tests and look for red flags like low GPA or test scores that do not meet the minimum requirements set by the university. Then I scan transcripts for low/failing grades on CS and related courses.  For these I make comments such as ``- unknown international school with good GPA, too many low grades on CS and related courses" or ``+ very good GPA from top school XYZ". I do not look at test details as the summary already gives the main scores.

% Then I look at LoRs.  I read carefully good LoRs and quickly skim over bad or generic ones (very easy to tell, you don't even need much experience to distinguish between strong LoRs from the rest). My comments for LoRs include ``+ very strong LoRs from well-known prof. XYZ talks about research experiences'' or ``- generic LoRs, mostly about course work''.  I also note any unusual or unique details in the LoRs, e.g., ``+ LoR mentioned about student's significant contribution on well-known open-source project X''. Sometimes a LoR came directly from a GMU professor who mentioned they already know the student and intend to offer RA, which is a very strong signal for admission (and my comment would be ``Prof. X making RA offer'').

% I skim over CV, looking for paper publications, research work experiences, and other relevant achievements such as winning prizes in well-known international programming/math competitions.  My notes for CV include things such as ``+ has X papers at top venues'' or ``+ gold medal at X''. Sometimes I also note about unusual or unique experiences, e.g., ``+ has a patent on X''.

% Next is the SOP, which I read carefully (but if it is bad then it would be a quick read). My notes for SOP include things such as ``- SOP mostly on course work, no clear motivation, generic" , or ``+ very enjoyable SOP talking about XYZ, customized for GMU, mentioned profs. X,Y,Z".  If the SOP mentioned publications then I would have ``+ has X papers at top venues" or ``+/- has X papers but at unknown places''.  I will not go online to search for the papers (unless they upload them as sample papers).  I also skim over sampling papers, essentially looking over the abstract to see if reads well, and make notes such as ``+ have a paper under submission to unknown venue, but reads well". If the student mentioned a project, I might also look at it (e.g., if it is on Github) and make notes such as ``+ has project on Github with X stars''.

% At this point I have enough to make up my decision, which I enter into the evaluation system. Typically my decisions are either \emph{reject} or \emph{admit with full funding}.  For the latter, I also mention funding sources (e.g., RA from prof. X or TA). I also include my comments to support the decision. Finally, I also recommend very strong students for the University Presidential fellowship (\autoref{sec:fellowships}).

% The adcom chair will look at all evaluations for each application and make a decision (likely following the decision of the reviewers).  If there are discrepancies such as one voting reject but another full funding acceptance (such cases are rare, but do happen), the chair might ask the reviewers to coordinate to reach a consensus.

>>>>

Many students lament that they get no interviews or are rejected (with strong profiles) and that the admission process seems random.  However, while it is true that the process is not perfect, it is not random.
There are many possible reasons for rejections!  Other than not having a strong profile (e.g., research potential, GPA, LoRs, SOP, poor interviews), here are some other reasons.

@@

\paragraph{Red flags} Various types of red flags can cause concern to the adcom (e.g., plagiarism, cheating, or other academic dishonesty). Another one is that you have a history of jumping from one program to another (e.g., you have been in multiple different programs in a short time). How would the reviewers know?  They might have contacts in other programs and heard about you, your LoRs might mention it.

@@

\section{How exactly do you evaluate an application?}\label{sec:ievaluate}

Around 15 minutes on average (less than 10 mins for definite rejection; 20+ mins for acceptance). In earlier years I spent more time, but now I can quickly tell if an application is good or bad.

The evaluation system allows for downloading a single pdf that consists of a summary page (e.g., schools, GPA, standard test scores), transcripts, test results, LoRs, CV, SOP, and writing samples (if any). I typically review the application in this order.  I also make notes on what I've seen, using them to support my evaluation.

I first skim the summary for school, GPA, standard tests and look for red flags like low GPA or test scores that do not meet the minimum requirements set by the university. Then I scan transcripts for low/failing grades on CS and related courses.  For these I make comments such as ``- unknown international school with good GPA, too many low grades on CS and related courses" or ``+ very good GPA from top school XYZ". I do not look at test details as the summary already gives the main scores.

Then I look at LoRs.  I read carefully good LoRs and quickly skim over bad or generic ones (very easy to tell, you don't even need much experience to distinguish between strong LoRs from the rest). My comments for LoRs include ``+ very strong LoRs from well-known prof. XYZ talks about research experiences'' or ``- generic LoRs, mostly about course work''.  I also note any unusual or unique details in the LoRs, e.g., ``+ LoR mentioned about student's significant contribution on well-known open-source project X''. Sometimes a LoR came directly from a GMU professor who mentioned they already know the student and intend to offer RA, which is a very strong signal for admission (and my comment would be ``Prof. X making RA offer'').

I skim over CV, looking for paper publications, research work experiences, and other relevant achievements such as winning prizes in well-known international programming/math competitions.  My notes for CV include things such as ``+ has X papers at top venues'' or ``+ gold medal at X''. Sometimes I also note about unusual or unique experiences, e.g., ``+ has a patent on X''.

Next is the SOP, which I read carefully (but if it is bad then it would be a quick read). My notes for SOP include things such as ``- SOP mostly on course work, no clear motivation, generic" , or ``+ very enjoyable SOP talking about XYZ, customized for GMU, mentioned profs. X,Y,Z".  If the SOP mentioned publications then I would have ``+ has X papers at top venues" or ``+/- has X papers but at unknown places''.  I will not go online to search for the papers (unless they upload them as sample papers).  I also skim over sampling papers, essentially looking over the abstract to see if reads well, and make notes such as ``+ have a paper under submission to unknown venue, but reads well". If the student mentioned a project, I might also look at it (e.g., if it is on Github) and make notes such as ``+ has project on Github with X stars''.

At this point I have enough to make up my decision, which I enter into the evaluation system. Typically my decisions are either \emph{reject} or \emph{admit with full funding}.  For the latter, I also mention funding sources (e.g., RA from prof. X or TA). I also include my comments to support the decision. Finally, I also recommend very strong students for the University Presidential fellowship (\autoref{sec:fellowships}).

The adcom chair will look at all evaluations for each application and make a decision (likely following the decision of the reviewers).  If there are discrepancies such as one voting reject but another full funding acceptance (such cases are rare, but do happen), the chair might ask the reviewers to coordinate to reach a consensus.
